---
title: "DS311 - R Lab Assignment"
author: "Isabella Barboza"
date: "11/6/2024"
output: html_document:
    theme: united
    highlight: tango
    df_print: paged
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Assignment 1

-   In this assignment, we are going to apply some of the build in data set in R for descriptive statistics analysis.
-   To earn full grade in this assignment, students need to complete the coding tasks for each question to get the result.
-   After finished all the questions, knit the document into HTML format for submission.

### Question 1

Using the **mtcars** data set in R, please answer the following questions.

```{r}
# Loading the data
data(mtcars)

# Head of the data set
head(mtcars)
```

a.  Report the number of variables and observations in the data set.

```{r}
# Enter your code here!
nrow(mtcars)
ncol(mtcars)
dim(mtcars)

# Answer:
cat("There are total of 32 variables and 11 observations in this data set.")
```

b.  Print the summary statistics of the data set and report how many discrete and continuous variables are in the data set.

```{r}
# Enter your code here!
data <- mtcars
summary(data)
discrete_vars <- sapply(data, function(x) 
  is.factor(x) ||
    is.character(x) || 
    is.logical(x))
continuous_vars <-  sapply(data, is.numeric)

num_discrete <- sum(discrete_vars)
cat(num_discrete)
num_continuous <- sum(continuous_vars)
cat(num_continuous)

# Answer:
print(paste("There are", num_discrete, "discrete variables and", num_continuous, "continuous variables in this data set."))
```

c.  Calculate the mean, variance, and standard deviation for the variable **mpg** and assign them into variable names m, v, and s. Report the results in the print statement.

```{r}
# Enter your code here!
m <- mean(mtcars$mpg)
v <- var(mtcars$mpg)
s <- sd(mtcars$mpg)

print(paste("The average of Mile Per Gallon from this data set is ",  m, " with variance ",  v, " and standard deviation",  s, "."))
```

d.  Create two tables to summarize 1) average mpg for each cylinder class and 2) the standard deviation of mpg for each gear class.

```{r}
# Enter your code here!
avg_mpg_cyl <- aggregate(mpg ~ cyl, data = mtcars, FUN = mean)
print(avg_mpg_cyl)

sd_mpg_gear <- aggregate(mpg ~ gear, data = mtcars, FUN = sd)
print(sd_mpg_gear)
```

e.  Create a crosstab that shows the number of observations belong to each cylinder and gear class combinations. The table should show how many observations given the car has 4 cylinders with 3 gears, 4 cylinders with 4 gears, etc. Report which combination is recorded in this data set and how many observations for this type of car.

```{r}
# Enter your code here!
crosstab <- table(mtcars$cyl, mtcars$gear)
print (crosstab)
max_combo <- which(crosstab == max(crosstab), arr.ind = TRUE)
print(max_combo)
print("The combination with the highest count is:")
  paste("Cylinders:", rownames(crosstab)[max_combo[1]])
  paste("Gears:", colnames(crosstab)[max_combo[2]])
  paste("Number of observations:", crosstab[max_combo])

  result <- paste("The most common car type in this data set is a car with", rownames(crosstab)[max_combo[1]], "cylinders and", colnames(crosstab)[max_combo[2]], "gears. There are a total of", crosstab[max_combo], "cars belonging to this specification in the data set.")
print(result)
```

### Question 2

Use different visualization tools to summarize the data sets in this question.

a.  Using the **PlantGrowth** data set, visualize and compare the weight of the plant in the three separated group. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your findings.

```{r}
# Load the data set
data("PlantGrowth")

# Head of the data set
head(PlantGrowth)

# Enter your code here!
library(ggplot2)

plantdf <- data.frame(
  weight = c(4.17, 5.58, 5.18, 6.11, 4.50, 4.61),
  group = factor(c("group1", "group2", "group3")) 
)
ggplot(plantdf, aes(x = group, y = weight)) + 
  geom_boxplot() +
  labs(
    title = "Plant Weights by Group",
    x = "Group",
    y = "Weight"
  )+
  theme_minimal()

#Summary

print("summary: Setting up the data and reviewing the data values from the data display, I noticed the graph missing assigned groups. I then labeled the data by separating plants into three groups and assigned it to x and labeled y as the weight of plant. I then opened a graphing package to plot the values that describe the connection between a group of plants and their weight.

By labeling X as the weight of the plant and Y as the group of plant with a range of 1:3, it displays a graph that shows the range and room for error within the weight for each group of plants. Group 1 has the widest range, while group 3 has the lowest range.")
```

Result:

=\> Report a paragraph to summarize your findings from the plot!

b.  Using the **mtcars** data set, plot the histogram for the column **mpg** with 10 breaks. Give labels to the title, x-axis, and y-axis on the graph. Report the most observed mpg class from the data set.

```{r}
library(ggplot2)

ggplot(mtcars, aes(x = mpg)) +
  geom_histogram(binwidth = 1, color = "black", fill = "pink") + 
  labs(
    title = "Observed Miles Per Gallon",
     x = "Miles Per Gallon (mpg)",
    y = "Frequency"  
  ) +
  theme_minimal()

mpg_table <- table(mtcars$mpg)  
most_observed_mpg <- names(mpg_table)[which.max(mpg_table)]  
most_observed_mpg
paste("Most of the cars in this data set are in the class of", most_observed_mpg, "mile per gallon.")
```

c.  Using the **USArrests** data set, create a pairs plot to display the correlations between the variables in the data set. Plot the scatter plot with **Murder** and **Assault**. Give labels to the title, x-axis, and y-axis on the graph. Write a paragraph to summarize your results from both plots.

```{r}
# Load the data set
data("USArrests")

# Head of the data set
head(USArrests)

# Enter your code here!
pairs(USArrests, 
      main = "USA Arrests Dataset",  
      pch = 19,  
      col = "pink",
      labs(
    title = "Observed Arrrests in the US",
    x = "Murder",
    y = "Assault",
    z = "Arrests"))



```

Result:

and =\> Report a paragraph to summarize your findings from the plot!

summary: Setting up the connection of the data and reviewing the data values from the data display for the different categories of arrests made are then applied to each state, which was grabbed using the heading data numbers. I plotted the US Arrests data by describing violent crimes committed within the graph. I assigned each value to a category, x as Murder, y as Assault, and z as Arrests. I opened a graphing package to plot the values that describe the connection between reason for arrest and assigned plotted points that represent each state and counts the reported arrests for murders and assaults and compares each category to one another, often times returning a positive correlation or close to no correlation.

------------------------------------------------------------------------

### Question 3

Download the housing data set from www.jaredlander.com and find out what explains the housing prices in New York City.

Note: Check your working directory to make sure that you can download the data into the data folder.

```{r, echo=FALSE}
# Check current working directory
 wd1 <- getwd()
 paste("Current Working Directory: ", wd1)

  target_dir <- "/Users/Bella/Desktop/DS_311/R"
if (dir.exists(target_dir)) {
   setwd(target_dir)
  wd2 <- getwd()
  cat("Current Working Directory: ", wd2)
} else {
  cat("Directory does not exist:", target_dir)
}
setwd("/Users/Bella/Desktop/DS_311/R")
wd2 <- getwd()
paste("Current Working Directory:", wd2)

# Load and clean the housing data set 
url <- 'http://www.jaredlander.com/data/housing.csv'
download.file(url = 'http://www.jaredlander.com/data/housing.csv', destfile = '~/Users/Bella/Desktop/DS_311/R/housing.csv', mode='wb')
housingData <- read.csv('/Users/Bella/Desktop/DS_311/R/housing.csv')
housingData <- subset(housingData,
                      select = c("Neighborhood", "Market.Value.per.SqFt", "Boro", "Year.Built")
housingData <- na.omit(housingData)                   
```

a.  Create your own descriptive statistics and aggregation tables to summarize the data set and find any meaningful results between different variables in the data set.

```{r}
data(housingData)
# Head of the cleaned data set
head(housingData)
nrow(housingData)
ncol(housingData)
dim(housingData)
# Enter your code here!
data <- housingData
summary(data)
discrete_vars <- sapply(data, function(x) 
  is.factor(x) ||
    is.character(x) || 
    is.logical(x))
continuous_vars <-  sapply(data, is.numeric)

num_discrete <- sum(discrete_vars)
cat(num_discrete)
num_continuous <- sum(continuous_vars)
cat(num_continuous)
print(paste("There are", num_discrete, "discrete variables and", num_continuous, "continuous variables in this data set."))

m <- mean(housingData$Market.Value.per.SqFt)
v <- var(housingData$Market.Value.per.SqFt)
s <- sd(housingData$Market.Value.per.SqFt)
#aggregate
avgMarVal_housSqFt_Neighbor <- aggregate(Market.Value.per.SqFt ~ Neighborhood, data = housingData, FUN = mean)
print(avgMarVal_housSqFt_Neighbor)
avgMarVal_housSqFt_year <- aggregate(Market.Value.per.SqFt ~ Year.Built, data = housingData, FUN = mean)
print(avgMarVal_housSqFt_year)
avgMarVal_housSqFt_buildingclass <- aggregate(Market.Value.per.SqFt ~ Building.Classification, data = housingData, FUN = mean)
print(avgMarVal_housSqFt_buildingclass)
avgMrVal_housSqFt_boro <- aggregate(Market.Value.per.SqFt ~ Boro, data = housingData, FUN = mean)
print(avgMarVal_housSqFt_buildingclass)
avgMrVal_housSqFt_GrossIncome <- aggregate(Market.Value.per.SqFt ~ Gross.Income.per.SqFt, data = housingData, FUN = mean)
print(avgMarVal_housSqFt_buildingclass)
# Aggregate mean Market Value per SqFt
 avghousingSqFt<- housingData %>%
  group_by(Neighborhood, Year.Built, Building.Classification, Boro, Gross.Income.per.SqFt) %>%
  summarise(avg_price_sqft = mean(Market.Value.per.SqFt, na.rm = TRUE))
print(agg_housingdata)


```

b.  Create multiple plots to demonstrates the correlations between different variables. Remember to label all axes and give title to each graph.

```{r}

library(ggplot2)

# Enter your code here!
data(housingData)
head(housingData)
numeric_data <- avghousingSqFt[, sapply(avghousingSqFt, is.numeric)]

pairs(numeric_data, 
      main = "Housing Dataset Correlations",
      pch = 19,               
      col = "pink",           
      labels = colnames(numeric_data),
      cex.labels = 0.8,       
      font.labels = 1)        

ggplot(avghousingSqFt, aes(x = Boro, y = Neighborhood)) +
  geom_point(color = "pink") +
  labs(
    title = "Average SqFt in Housing",
    x = "Boro",
    y = "Neighborhood"
  )

```

c.  Write a summary about your findings from this exercise.

=\> Enter your answer here!

From aggregating the data and creating a new data set we can see a correlation between gross income per square foot and average price per sqaure feet form a positive correlation. With inclusion of year built, we see an increase in homes after the 1900's. When comparing Neighborhoods and Boroughs to average housing square feet, there is more housing within Brooklyn than other boroughs in New York.
